Determination of the \ce{^{64}Zn},\ce{^{47}Ti}(n,p) Cross-Sections using a DD Neutron Generator for Medical Isotope Studies

### 2017-02-24 First_np_Paper/first_np_paper.tex
Updated introduction section with Lee edits

Cross sections for the \ce{^{47}Ti}(n,p)\ce{^{47}Sc} and \ce{^{64}Zn}(n,p)\ce{^{64}Cu} reactions have been measured for quasi-monoenergetic DD neutrons produced by the UC Berkeley High Flux Neutron Generator. The study was motivated by interest in the production of \ce{^{47}Sc} and \ce{^{64}Cu} as emerging medical isotopes. The cross sections were measured in ratio to the \ce{^{113}In}(n,n')\ce{^{113m}In} and \ce{^{115}In}(n,n')\ce{^{115m}In} inelastic scattering reactions on co-irradiated indium samples. Post-irradiation counting using an HPGe and LEPS detectors allowed for cross section determination to within 6\% uncertainty. The cross section were determined with lower uncertainty than existing measurements and are found to be  in good agreement with the literature and theoretical values. This work highlights the utility of using D-D plasma-based neutron sources, for a host of nuclear data measurements and potentially for the production of radionuclides for medical applications.

@@

\comment{Add an initial \enquote{punchline} to  clarify the motivation?}

\comment{Per Mauricio, does 2.7 MeV count as high energy? While fast, I would consider this to still be definitely in the low-energy regime, compared to spallation, etc.}

A series of experiments are being conducted at UC Berkeley to measure low-energy neutron-induced production cross sections for a range of emerging medical radioisotopes, as many of these cross sections currently have not been measured, or possess significant uncertainties.   One potential method for producing many of these isotopes is the use of (n,p) reactions via neutron beams from DD neutron generators. Direct neutron activation, using a DD neutron generator, allows for more precise cross section measurements than time-of-flight methods, and potentially offers a proliferation-resistant pathway for radioisotope production. Neutron activation can provide several advantages over charged-particle reactions from a targetry perspective. Notably, the higher penetration of neutrons allows for the use of thick target designs, which can improve production throughput and greatly simplifies target cooling considerations. More importantly, low-energy neutron reactions can be used to prevent competing reaction channels in a charged-particle reaction which would otherwise contaminate the product specific activity.

The present work focuses on the (n,p) production cross section measurements of \ce{^{64}Cu} and \ce{^{47}Sc}. Copper-64  ($t_{1/2}$ = 12.7 h) undergoes $\beta^+$ decay (61.5\% branching ratio) to \ce{^{64}Ni} or $\beta^-$ decay (38.5\% branching ratio) to \ce{^{64}Zn}. The emitted 190-keV $\beta^-$ particle makes this an  attractive short-ranged therapeutic radionuclide, which also has the possibility for simultaneous positron emission tomography (PET) imaging  due to the positron emission branch (for real-time dose monitoring and verification),  making \ce{^{64}Cu} particularly attractive for emerging radiation therapy protocols. In addition, copper radiochemistry is well developed, and many existing ligands and carriers may be used for selective delivery of the radionuclide to different sites in patients.

Scandium-47 ($t_{1/2}$ = 3.35 d) undergoes $\beta^-$ decay to \ce{^{47}Ti}, emitting a high-intensity 159-keV gamma ray in the process. This radionuclide is  attractive as an emerging diagnostic isotope, due to the similarity of the emitted gamma ray to that of the more well-established \ce{^{99m}Tc}. However, due to the short half-life  and dwindling supplies of \ce{^{99m}Tc}, \ce{^{47}Sc} stands poised as a potential solution to this shortage, due to its longer half-life and multiple production pathways without the need for highly enriched uranium.

\comment{Adding a short paragraph from Mauricio explaining benefits of DD generators vs other neutron sources?}

\comment{Per Daniel, add a short summary of results here?}

Current methodology in radiochemistry has been proven to successfully recover upwards of 95\% of produced \ce{^{64}Cu} and \ce{^{47}Ti} from solid target designs, without the need for additional carrier. By expanding the base of efficient reaction pathways, great advances can be made in making production of medical radionuclides more efficient and affordable for those in need.

\comment{Should this last paragraph be moved to the results and discussion at the end of the paper?}

@@

\begin{figure*}
    \centering
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
%         \includegraphics[width=\columnwidth]{./figures/Capture.PNG}
        \includegraphics[height=3in]{./figures/Capture.PNG}
        \caption{The UC Berkeley High-Flux Neutron Generator,}
        \label{fig:hfng_a}
    \end{subfigure}%
    ~ 
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
%         \includegraphics[width=\textwidth]{./figures/target2.png}
        \includegraphics[height=2in]{./figures/target2.png}
        \caption{Cutaway of the target chamber.}
                \label{fig:hfng_b}
    \end{subfigure}
    \caption{Note: the ion source is approximately 20 cm in diameter.}
     \label{fig:main_HFNG}
\end{figure*}

>>>>

Cross sections for the \ce{^{47}Ti}(n,p)\ce{^{47}Sc} and \ce{^{64}Zn}(n,p)\ce{^{64}Cu} reactions have been measured for quasi-monoenergetic DD neutrons produced by the UC-Berkeley High Flux Neutron Generator. The study was motivated by interest in the production of \ce{^{47}Sc} and \ce{^{64}Cu} as emerging medical isotopes. The cross sections were measured in ratio to the \ce{^{113}In}(n,n')\ce{^{113m}In} and \ce{^{115}In}(n,n')\ce{^{115m}In} inelastic scattering reactions on co-irradiated indium samples. Post-irradiation counting using an HPGe and LEPS detectors allowed for cross section determination to within 6\% uncertainty. The cross section were determined with lower uncertainty than existing measurements and are found to be  in good agreement with the literature and theoretical values. This work highlights the utility of using DD plasma-based neutron sources, for a host of nuclear data measurements and potentially for the production of radionuclides for medical applications.

@@

There has been significant interest in the past several years in exploring the use of neutron-induced reactions to create radionuclides for a wide range of applications, due to the far greater range of neutrons as compared to charged particle beams ($\sfrac{\text{g}}{\text{cm}^2}$ as compared to 10's of $\sfrac{\text{mg}}{\text{cm}^2}$), together with the fact that isotope production facilities often produce large secondary neutron fields.  Particular interest has been paid to (n,p) and (n,$\alpha$) charge-exchange reactions since these reactions produce high-specific activity radionuclide samples without the use of chemical carriers in the separation process.

Two other potential neutron sources for (n,x) reactions exist in addition to the secondary neutron fields generated at existing isotope production facilities: reactors and neutron generators that utilize the D(T,n)$\alpha$ \enquote{DT} and D(D,n)\ce{^3He} \enquote{DD} reactions.  While reactors produce copious quantities of neutrons, their energy spectrum is often not well-suited to the preparation of high-purity samples due to the co-production of unwanted activities via neutron capture, in addition to the significant start-up costs and proliferation concerns involved in their commissioning.  Similarly, while the higher energy 14-15 MeV neutrons produced at DT generators are capable of initiating (n,p) and (n,$\alpha$) reactions, their higher energy opens the possibility of creating unwanted activities via (n,pxn) and (n,$\alpha$xn) reactions that cannot easily be separated from the desired radionuclides.

In contrast, while the neutron spectrum from a DD reaction, which ranges from approximately 2-3 MeV, is ideally suited to radionuclide production, the lower flux from these generators limits their production rate.  An additional complication is the relative paucity of high-quality, consistent cross section data for neutrons in the 2-3 MeV DD energy range.

The purpose of the present work is to explore the potential for using high-flux neutron generators to make high-specific activity samples of radionuclides at the mCi level for local use in the application community.  The research group at UC-Berkeley has been developing a High Flux Neutron Generator (HFNG) that features an internal target where samples can be placed several mm from the neutron producing surface in order to maximize the utilization of each neutron for the production of a desired radionuclide \textcolor{red}{[XXXref Cory paper]}.  The HFNG uses the D(D,n)\ce{^3He} reaction to produce neutrons with energies near 2.45 MeV together with a self-loading target design to maintain continuous operation without target replacement.  In addition to the generator itself, efforts are underway to design neutron reflection capabilities to allow scattered neutrons multiple opportunities to interact with an  internally mounted target. While these design efforts are underway, the HFNG can be used to better characterize production cross sections at the appropriate neutron energy.

\comment{cite Cory's paper}

The present work features a pair of cross section measurements for the production of two non-standard positron emitters: \ce{^{64}Zn}(n,p)\ce{^{64}Cu} and \ce{^{47}Ti}(n,p)\ce{^{47}Sc}. \ce{^{64}Cu}  ($t_{1/2}$ = 12.7 h) undergoes $\beta^+$ decay (61.5\% branching ratio) to \ce{^{64}Ni} or $\beta^-$ decay (38.5\% branching ratio) to \ce{^{64}Zn}. The emitted 190-keV $\beta^-$ particle makes this an  attractive short-range therapeutic radionuclide, which also has the possibility for simultaneous positron emission tomography (PET) imaging for real-time dose monitoring and verification. This makes \ce{^{64}Cu} particularly attractive for emerging radiation therapy protocols. In addition, copper radiochemistry is well developed, and many existing ligands and carriers may be used for selective delivery of the radionuclide to different sites in patients. The second radionuclide studied, \ce{^{47}Sc} ($t_{1/2}$ = 3.35 d), undergoes $\beta^-$ decay to \ce{^{47}Ti}, emitting a high-intensity 159-keV gamma ray in the process. This radionuclide is  attractive as an emerging diagnostic isotope, due to the similarity of the emitted gamma ray to that of the more well-established \ce{^{99m}Tc}. However, due to the short half-life of and dwindling supplies of \ce{^{99m}Tc}, \ce{^{47}Sc} stands poised as a potential solution to this shortage, due to its longer half-life and multiple production pathways without the need for highly enriched uranium. In addition, when paired with \ce{^{44}Sc}, \ce{^{47}Sc} forms a promising \enquote{theranostic} set for use in simultaneous therapeutic and diagnostic applications.

\comment{cite theranostic papers, etc}

Current methodology in radiochemistry has been proven to successfully recover upwards of 95\% of produced \ce{^{64}Cu} and \ce{^{47}Sc} from solid target designs, without the need for additional carrier. By expanding the base of efficient reaction pathways, great advances can be made in making production of medical radionuclides more efficient and affordable for those in need.

@@

\begin{figure*}
    \centering
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
%         \includegraphics[width=\columnwidth]{./figures/Capture.PNG}
        \includegraphics[height=3in]{./figures/Capture.PNG}
        \caption{The UC-Berkeley High-Flux Neutron Generator,}
        \label{fig:hfng_a}
    \end{subfigure}%
    ~ 
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
%         \includegraphics[width=\textwidth]{./figures/target2.png}
        \includegraphics[height=2in]{./figures/target2.png}
        \caption{Cutaway of the target chamber.}
                \label{fig:hfng_b}
    \end{subfigure}
    \caption{Note: the ion source is approximately 20 cm in diameter.}
     \label{fig:main_HFNG}
\end{figure*}
